Self Referencing Sequences

### 2014-11-06 draft.tex
rewrote densities section, broke it up into two sections

\section{Density}
\label{density}
\emph{Written by Erik Waingarten, edited by Perry Kleinhenz and Fermi Ma}

A topic of interest when considering a self referencing sequence is the relative density of the various elements of the generating cycle. It is unclear if such a density even exists and if so how to compute it for arbitrary generating cycles. In this section, we show how to use substitution rules to compute relative densities for certain sequences if we assume such densities exists. We explain the method and compute the relative density of 3's for $C = (1 3)$. We will also produce bounds on the density of $2$'s $C = (1 2)$ in a way that suggests that the density converges to 2; however, this method is computationally difficult for smaller bounds since the number of cases to check increases exponentially.

To begin we will discuss a technique that can be used to compute the densities for some generating cycles. In particular, suppose we have a generating cycle $C$ with a set of rules $R$ that satisfy the conditions from the section above. For each rule $r \in R$, we will say denote the rule as $r_l \rightarrow r_r$ as the inputs and the outputs.

Now we can count the fraction of times $r_r$ appears in the sequence at each step. Since we know the rules, these will become a set of $|R|$ equalities that reference each other. We can take their ratios and solve for the fraction of times that each $r_r$ appears and compute the density by assuming a stable state.

For example, lets look at the rules we have determined for $G = \{ 1, 3 \}$. 
\begin{align*}
\text{Rule }A: 3,3 &\to 3,3,3,1,1,1 \\
\text{Rule }B: 3,1 &\to 3,3,3,1\\
\text{Rule }C: 1,1 &\to 3,1
\end{align*}
Then we can let $3,3,3,1,1,1 = A$, $3,3,3,1 = B$ and $3, 1 = C$. If $A_i$ is the number of occurrences of $A$ in $s^{(i)}$, then 
\[ A_i = B_{i-1} + A_{i-1} \]
Likewise, we have
\[ B_i = A_{i-1} + B_{i-1} + C_{i-1} \]
\[ C_i = A_{i-1} \]
So now we can take the fraction by dividing by the total number of them:
\[ A_i' = \dfrac{B_{i-1}' + A_{i-1}'}{3A_{i-1}' + 2B_{i-1}' + C_{i-1}'} \]
\[ B_i' = \dfrac{A_{i-1}' + B_{i-1}' + C_{i-1}'}{3A_{i-1}' + 2B_{i-1}' + C_{i-1}'} \]
\[ C_i' = \dfrac{A_{i-1}'}{3A_{i-1}' + 2B_{i-1}' + C_{i-1}'} \]
So if we assume a density, there will be a stable state which will be the density, so in the case where $A_i'=A_{i-1}', B_i' = B_{i-1}', C_i'=C_{i-1}'$, we can solve for the fractional appearance of each and get the density of the numbers.

We can compute solutions to this set of quadratic equations and find the density of 3's is approximately $0.602$.

As discussed in the previous section, it is unclear how to establish a set of rules for $C = (1 2)$; however, there is some structure that we can exploit. In particular, we know that the sequences $1, 1, 1$ and $2, 2, 2$ will never appear in the self-referential sequence. This is because such sequences would be part of a block of size at least $3$.

>>>>

\section{A Density Computation}
\label{densitycomputation}
\emph{Written by Erik Waingarten and Fermi Ma, edited by Perry Kleinhenz}

An open question regarding the self-referential sequence of 1's and 2's, $s((1 2),1)$, is what the limiting \emph{density} of 1's and 2's is. In this section, we consider the related problem of finding the density for sequences with generating cycle $C = (1 3)$ (note that the sequence that starts with a 1 differs by only one number from the sequence that starts with a 3, so we do not need to consider them separately). While we cannot prove that a limiting density exists, we give a numerical calculation of what the density is \emph{if it exists}. We use a general technique that takes advantage of the fact that substitution rules when $C = (1 3)$, as shown in Section~\ref{substitutionrules}.

In Section~\ref{substitutionrules}, we found that repeated applications of the rules
\begin{align*}
\text{Rule }1: 3,3 &\to 3,3,3,1,1,1 \\
\text{Rule }2: 3,1 &\to 3,3,3,1\\
\text{Rule }3: 1,1 &\to 3,1
\end{align*}
to the initial sequence $\{3,1\}$ gives a growing sequence that converges to the self-referential sequence $s((1 3),3)$.

If we let $A$ denote the subsequence $\{3,3\}$, $B$ denote $\{3,1\}$ and let $C$ denote $\{1,1\}$, then the rules can be rewritten as
\begin{align*}
\text{Rule }1: A &\to ABC \\
\text{Rule }2: B &\to AB\\
\text{Rule }3: C &\to B,
\end{align*}
where AB and ABC are to be interpreted as concatenations of the subsequences.

Our technique for computing densities will work as follows. Let $N_i(A), N_i(B),$ and $N_i(C)$ denote the number of $A$'s, $B$'s, and $C$'s, respectively, in the sequence given by the $i$th application of the substitution rules. We keep track of these counts of $A$'s, $B$'s, and $C$'s at each round of rule applications, and we use these to determine limiting proportions of $A$'s, $B$'s and $C$'s. From these limiting proportions, we can deduce the densities of 1's and 3's.

To start, consider the substitution process from the very first step. We start applying substitution rules to the initial sequence $\{3,1\}$. Since $\{3,1\} = B$, the initial counts are:
\[\text{Sequence at step }1: \{3,1\},\text{ Counts}: N_1(A) = 0, N_1(B) = 1, N_1(C) = 0.\]

After one application of the substitution rules, we have
\[\text{Sequence at step }2: \{3,3,3,1\},\text{ Counts}: N_2(A) = 1, N_2(B) = 1, N_2(C) = 0.\]

After one more application, we have
\[\text{Sequence at step }3: \{3,3,3,1,1,1,3,3,3,1\}, \text{ Counts}: N_3(A) = 2, N_3(B) = 2, N_3(C) = 1.\]

For the purposes of computing densities, it turns out that only the counts are relevant and we do not need to keep track of the sequence at each step. We can compute the counts $N_i$ from the values of $N_{i-1}$ as follows
\begin{align*}
N_i(A) &= N_{i-1}(A) + N_{i-1}(B)\\
N_i(B) &= N_{i-1}(A) + N_{i-1}(B) + N_{i-1}(C)\\
N_i(C) &= N_{i-1}(A).
\end{align*}
The above equations are arrived at by looking at the set of rules involving $A,B,$ and $C$. For example, an $A$ at the $i$th iteration is written down every time there is an $A$ or $B$ in the previous iteration, which leads to the first equation above.

For computing densities, however, the \emph{proportions} of $A$'s, $B$'s, and $C$'s are needed. Fortunately, we can derive equations for the proportions by simply normalizing the equations for the counts. Let $P_i(X)$ denote the proportion of the sequence after iteration $i$ that is made up of the $X$ subsequence. Normalizing gives the following equations:
\begin{align*}
P_i(A) &= \frac{P_{i-1}(A) + P_{i-1}(B)}{3P_{i-1}(A) + 2P_{i-1}(B) + P_{i-1}(C)}\\
P_i(B) &= \frac{P_{i-1}(A) + P_{i-1}(B) + P_{i-1}(C)}{3P_{i-1}(A) + 2P_{i-1}(B) + P_{i-1}(C)}\\
P_i(C) &= \frac{P_{i-1}(A)}{3P_{i-1}(A) + 2P_{i-1}(B) + P_{i-1}(C)}.
\end{align*}
The only difference is that each equation is divided on the right-hand side by $3P_{i-1}(A) + 2P_{i-1}(B) + P_{i-1}(C)$ to ensure that the proportions of $A$'s, $B$'s, and $C'$s at each step add up to 1. Now, assume that there exist limiting proportions (and hence, limiting densities) as $i \to \infty$. Then at this limit, $P_i(X) = P_{i-1}(X) = P_{\infty}(X)$. Making these substitutions and solving for $P_{\infty}(A), P_{\infty}(B),$ and $P_{\infty}(C)$ in \texttt{Mathematica} gives the following values:
\begin{align*}
P_{\infty}(A) &\approx \texttt{0.3760858894420931}\\
P_{\infty}(B) &\approx \texttt{0.4533976515164037}\\
P_{\infty}(C) &\approx \texttt{0.1705164590415032}.
\end{align*}

Since $A = \{3,3\}$, $B = \{3,1\}$ and $C = \{1,1\}$, the fraction of 1's in the limiting sequence is $0.5P_{\infty}(B) + P_{\infty}(C)$, and the fraction of 3's in the limiting sequence is $P_{\infty}(A) + 0.5P_{\infty}(B)$. So the density values, if they exist, are
\begin{align*}
&\text{Fraction of 1's: }\texttt{0.397215284799705}\\
&\text{Fraction of 3's: }\texttt{0.602784715200295}.
\end{align*}
We were able to support these calcuations with numerical simulations. We generated the $s((13),3)$ sequence up to 100,000 terms, and we found that the fraction of 1's was approximately $0.398$, and that the fraction of 3's was approximately 0.602.

It is difficult to fully generalize this procedure for all substitution rule sets, as we cannot prescribe a general form for all substitution rules. However, we believe that this technique should work whenever all the inputs to the substitution rules are the same length (in this case, the inputs were all length 2), and the outputs are all lengths that are multiples of the inputs. We do not know if the resulting equations will always give exactly 1 solution, as they did in this case.

\section{Density Bounds for $C = (12)$}
\label{densitybounds}

As discussed in the Section~\ref{substitutionrules}, it is unclear how to establish a set of rules for $C = (1 2)$; however, there is some structure that we can exploit. In particular, we know that the sequences $1, 1, 1$ and $2, 2, 2$ will never appear in the self-referential sequence. This is because such sequences would be part of a block of size at least $3$.
